---
title: "Ovarian cancer survival analisys"
author: ""
date: "`r format(Sys.Date(), format = '%d.%m.%Y')`"
output:
    html_document:
      toc: true
      toc_depth: 2
      toc_float: true
      number_section: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

This report represents survival analysis using ovarian cancer data set from package 'survival' and includes short EDA, Kaplan-Meier survival curves and assessment of the factors that influence on survival.

# Methods

All computations were performed in RStudio with R version 3.6.3. Required packages:
```{r installing packages, message=FALSE}
require(survival)
require(survminer)
require(ggplot2)
require(coin)
require(ggfortify)
require(ggpubr)
```

# EDA

```{r data, message=FALSE, warning=FALSE}
ov <- ovarian
str(ov)
summary(ov)
```

Variables:

*futime* - survival or censoring time

*fustat* - censoring status

*age* - in years

*resid.ds* - residual disease present (1=no, 2=yes)

*rx* - treatment group

*ecog.ps* - ECOG performance status - functional status of a patient (1 is better)

```{r NA, message=FALSE, warning=FALSE}
sum(is.na(ov))
```

Conversion into factors:

```{r to factors, message=FALSE, warning=FALSE}
ov$resid.ds <- factor(ov$resid.ds, labels = c('No', 'Yes'))
ov$rx <- factor(ov$rx, labels = c('Endoxan', 'Endoxan+Adria'))
ov$ecog.ps <- factor(ov$ecog.ps, labels = c('better', 'worse'))
```

```{r distribution, message=FALSE, warning=FALSE}
age_plot <- ggplot(ov, aes(age))+
  geom_histogram(fill = 'cadetblue3', binwidth = 5, col = 'black')+
  theme_bw()+
  labs(title = "Distribution \nof the patient age")+
  theme(plot.title = element_text(size=10))
futime_plot <- ggplot(ov, aes(futime))+
  geom_histogram(fill = 'cadetblue3', binwidth = 80, col = 'black')+
  theme_bw()+
  labs(title = "Distribution \nof the survival time")+
  theme(plot.title = element_text(size=10))
fustat_plot <- ggplot(ov, aes(factor(fustat, labels =  c('Censored', 'Death'))))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "Censoring status")+
  xlab('fustat')+
  theme(plot.title = element_text(size=10))
resid.ds_plot <- ggplot(ov, aes(resid.ds))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "Residual disease presence")+
  xlab('resid.ds')+
  theme(plot.title = element_text(size=10))
rx_plot <- ggplot(ov, aes(rx))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "Treatment")+
  xlab('rx')+
  theme(plot.title = element_text(size=10))
ecog_plot <- ggplot(ov, aes(ecog.ps))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "ECOG perfomance status")+
  xlab('ecog.ps')+
  theme(plot.title = element_text(size=10))
arr <- ggarrange(age_plot, futime_plot, fustat_plot, resid.ds_plot, rx_plot, ecog_plot, ncol=3, nrow=2)
arr
```

It makes sense to create a new dummy variable of age. The median value of age was taken as the separation basis.

```{r dummy_age, message=FALSE, warning=FALSE}
ov$dummy_age <- ifelse(ov$age > 57, 'Older', 'Younger')
ov$dummy_age <- as.factor(ov$dummy_age)
dummy_age_plot <- ggplot(ov, aes(dummy_age))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "Age of the patients")+
  xlab('dummy_age')
dummy_age_plot
```

To assess the differences between groups of different age and treatment a hybrid variable was created:

```{r treatment_age, message=FALSE, warning=FALSE}
ov$rx_age <- interaction(ov$rx, ov$dummy_age)
rx_age_plot <- ggplot(ov, aes(rx_age))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "Age of the patients and treatment")+
  xlab('rx_age')
rx_age_plot
```

To assess the differences between groups of different age and ECOG status a hybrid variable was created:

```{r ecog_age, message=FALSE, warning=FALSE}
ov$ecog_age <- interaction(ov$ecog.ps, ov$dummy_age)
ecog_age_plot <- ggplot(ov, aes(ecog_age))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "Age of the patients and ECOG status")+
  xlab('ecog_age')
ecog_age_plot
```

To assess the differences between groups of different ECOG status and treatment a hybrid variable was created:

```{r treatment_ecog, message=FALSE, warning=FALSE}
ov$rx_ecog <- interaction(ov$rx, ov$ecog.ps)
rx_ecog_plot <- ggplot(ov, aes(rx_ecog))+
  geom_bar(fill = 'cadetblue3', col = 'black', width = 0.5)+
  theme_bw()+
  labs(title = "ECOG status of the patients and treatment")+
  xlab('rx_age')
rx_ecog_plot
```

*Summarizing of the EDA results:*

- most variables are nominative;

- no NA's;

- age of the patients distributed almost normally;

- death or censored time distribution is not normal (or normal by a stretch of imagination);

- slightly more than half of the patients is censored;

- more patients had residual disease after treatment;

- treatment groups were balanced, with equal quantity;

- slightly more patients had better ECOG performance status of 1, others had status 2;

- groups by age and treatment are not well balanced;

- groups by age and ECOG status are almost well balanced;

- groups by treatment and ECOG status are almost well balanced.


# Kaplan-Meier Survival Curves

```{r km, message=FALSE, warning=FALSE}
km <- with(ov, Surv(futime, fustat))
km
```

```{r survival curve, message=FALSE, warning=FALSE}
km_fit <- survfit(km ~ 1, data=ov)
summary(km_fit)
autoplot(km_fit, main = 'Survival curve')+
  theme_bw()
```

From the Kaplan-Meier curve we can see that the experiment had finished with censored patients.

# Differences in survival between groups

## Influence of different factors

Influence of the treatment on survival:

```{r km_treatment, message=FALSE, warning=FALSE}
km_treatment_fit <- survfit(km ~ rx, data=ov)
autoplot(km_treatment_fit, main = 'Survival curves, by treatment')+
  theme_bw()
```

Influence of the age on survival:

```{r km_age, message=FALSE, warning=FALSE}
km_age_fit <- survfit(km ~ dummy_age, data=ov)
autoplot(km_age_fit, main = 'Survival curves, by age')+
  theme_bw()
```

Influence of the residual disease presence on survival:

```{r km_resid, message=FALSE, warning=FALSE}
km_resid_fit <- survfit(km ~ resid.ds, data=ov)
autoplot(km_resid_fit, main = 'Survival curves, by residual disease presence')+
  theme_bw()
```

Influence of the ECOG performance status on survival:

```{r km_ecog, message=FALSE, warning=FALSE}
km_ecog_fit <- survfit(km ~ ecog.ps, data=ov)
autoplot(km_ecog_fit, main = 'Survival curves, by ECOG status')+
  theme_bw()
```
Influence of the age and treatment on survival:

```{r km_treatment_age, message=FALSE, warning=FALSE}
km_rx_age_fit <- survfit(km ~ rx_age, data=ov)
autoplot(km_rx_age_fit, main = 'Survival curves, by treatment and age')+
  theme_bw()
```

Influence of the age and ECOG status on survival:

```{r km_ecog_age, message=FALSE, warning=FALSE}
km_ecog_age_fit <- survfit(km ~ ecog_age, data=ov)
autoplot(km_ecog_age_fit, main = 'Survival curves, by ECOG status and age')+
  theme_bw()
```

Influence of the ECOG status and treatment on survival:

```{r km_treatment_ecog, message=FALSE, warning=FALSE}
km_rx_ecog_fit <- survfit(km ~ rx_ecog, data=ov)
autoplot(km_rx_ecog_fit, main = 'Survival curves, by treatment and ECOG status')+
  theme_bw()
```

*Results of survival curves' analysis:* (only visual analysis!)

- it seems that Endoxan+Adria treatment worked better than Endoxan only;

- younger patients survived better, but many of them were censored on the segment [750-1200];

- patients with no residual disease survived better;

- patients with better ECOG perfomance status survived better on the segment [500-1200];

- older patients treated with Endoxan only survived worse than treated with Endoxan+Adria, and worse than younger patients regardless of the treatment of the latter;

- no obvious difference between younger patients with different treatment;

- logically, older patients with worse/better status survived worse than younger with worse/better status;

- patients with better status treated with Endoxan+Adria survived better then others, with six censored patients (maybe recovered?).

## Log-rank tests

Log-rank tests were performed to assess the significance of the differences between groups.

```{r lr_treatment, message=FALSE, warning=FALSE}
survdiff(formula = km ~ rx, data = ov)
logrank_test(km ~ rx, data = ov)
```

```{r lr_age, message=FALSE, warning=FALSE}
survdiff(formula = km ~ dummy_age, data = ov)
logrank_test(km ~ dummy_age, data = ov)
```

```{r lr_resid, message=FALSE, warning=FALSE}
survdiff(formula = km ~ resid.ds, data = ov)
logrank_test(km ~ resid.ds, data = ov)
```

```{r lr_ecog, message=FALSE, warning=FALSE}
survdiff(formula = km ~ ecog.ps, data = ov)
logrank_test(km ~ ecog.ps, data = ov)
```


```{r lr_treatment_age, message=FALSE, warning=FALSE}
survdiff(formula = km ~ rx_age, data = ov)
logrank_test(km ~ rx_age, data = ov)
```

```{r lr_ecog_age, message=FALSE, warning=FALSE}
survdiff(formula = km ~ ecog_age, data = ov)
logrank_test(km ~ ecog_age, data = ov)
```

```{r lr_treatment_ecog, message=FALSE, warning=FALSE}
survdiff(formula = km ~ rx_ecog, data = ov)
logrank_test(km ~ rx_ecog, data = ov)
```

*Results of the group differences analysis:*

- no difference significant differences between treatment groups (Z=-1.0296, p-value=0.3032);

- no differences between age groups (Z=-1.5717, p-value=0.116);

- almost significant difference between patients with and without residual disease (Z=1.9234, p-value=0.05444);

- no difference between ECOG groups (Z=0.69106, p-value=0.4895);

- no difference between four groups by age and treatment (chi-squared = 5.4522, df = 3, p-value = 0.1415);

- no difference between four groups by age and ECOG status (chi-squared = 2.8655, df = 3, p-value = 0.4128);

- no difference between four groups by treatment and ECOG status (chi-squared = 3.159, df = 3, p-value = 0.3677).


# Analysis of the factors which affect the survival

## Cox model

```{r cox, message= FALSE, warning=FALSE}
cox <- coxph(formula = km ~ age + resid.ds + rx + ecog.ps , data = ov)
summary(cox)

cox_fit <- survfit(cox)

autoplot(cox_fit, main = 'Kaplan-Meier survival curve, corrected')+
  theme_bw()
```

According to the Cox model only age variable is significant (p-value=0.00777), but covariates in this model are non variable during time. Changing of each component is shown on the plots below.

```{r covariates, message=FALSE, warning=FALSE}
aa_fit <- aareg(km ~ age + resid.ds + rx + ecog.ps , data = ov)
autoplot(aa_fit, main = 'Changing of the model components')+
  theme_bw()
```

## Risk ratio

```{r risk ratio, message=FALSE, warning=FALSE}
fit.coxph <- coxph(km ~ dummy_age + resid.ds + rx + ecog.ps , data = ov)
ggforest(fit.coxph, data = ov)
```

Risk ratio doesn't significantly change between groups of patients. P-values represented on the plot above are lower then 0.05, but we can see some tendencies. Perhaps the larger group of patients would provide more relevant results.








